When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-02-28 doc/paper/vpd_et_paper.tex
done spell check

The opposing effects of increased atmospheric demand and higher stomatal stress lead to two possible perspectives for how evapotraspiration (ET) respondes to shifts in VPD. The first, a hydrometeorological perspective, is that higher VPD increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indices and hydrometeorological studies \citep[e.g.,][]{Heim_2002, Scheff_2015}, typically only quantifies changes in atmospheric demand and fails to account for ecosystem response \citep{Swann_2016}. In reality, plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to partially close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. This leads to a plant physiology perspective, in which an increase in VPD, particularly in well-watered soil conditions, may actually correspond to a decrease in ET because of stomatal closure. This decrease in ET would reduce and potentially cancel, in the case of full stomatal closure, the effects of shifts to atmospheric demand. In other words, the  question ``When does VPD drive or reduce ET?'' can be related to whether plant regulation or atmospheric demand dominates ET response.

@@

\section{Materials and Methods}
\label{methods}
\subsection{Data}
\label{data}
We use both meteorological and eddy-covariance data from the FLUXNET2015 database, including all sites with at least four years of data (data available at \sloppy https://fluxnet.fluxdata.org/data/fluxnet2015-dataset/ \sloppy). Each site's plant functional type (PFT) was classified using the International Geosphere-Biosphere Programme vegetation classification scheme \citep{Loveland_1999}. The physical constants used in the stomatal conductance model in the Methods section (Section \ref{methods}) are only published for six plant functional types (PFTs): crops (CRO), deciduous broadleaf forest (DBF), evergreed broadleaf forests (EBF), evergreen needleleaf forest (ENF), grass (GRA), and closed shrub (CSH) (see Table \ref{pft}). There are 62 sites with these plant functional types, and their location is shown in  Figure \ref{map_fig}.

\begin{table}
  \caption{Plant functional types, their abbreviation, Medlyn coefficient (from \citet{Lin_2015}, via \citet{Franks_2017}) and caluclated uWUE. uWUE from \citet{Zhou_2015}, including the  observed standard deviation is shown for comparison. Note that uWUE from \citet{Zhou_2015} is calculated from a different set of sites, and that units are converted such that the quantities fit into Equations 1-8 with the variables in Table \ref{definitions}.}
    \small
\label{pft}
\centering
\begin{tabular}{l c c @{\qquad} c c}
  \hline
  \multirow{2}[3]{*}{Abbreviation} & \multirow{2}[3]{*}{PFT} & \multirow{2}[3]{*}{$g_1$ (Pa$^{0.5}$)} & \multicolumn{2}{c}{uWUE ($\mu$-mol [C] Pa$^{0.5}$ J$^{-1}$ [ET])}  \\
  \cmidrule{4-5}

@@

Differences in $VPD_{crit}$ are exclusively determined by uWUE and the slope parameter ($g_1$) related to the plant functional type. A larger uWUE means a smaller $VPD_{crit}$, and an ET response to increases VPD that is more likely to be positive. At first glance this result is somewhat counterintuitive; we expect that plants with a higher water use effiency would be more water conservative. However, in reality uWUE determines how $WUE$ changes with VPD:

@@

So, plants with a higher uWUE will have a greater \textit{decrease} in ecosystem-scale $WUE$ in response to increases in VPD. This decrease in $WUE$ causes more water loss per unit carbon production, and explains the relationship between high uWUE and high liklihood of increases of ET in response to increasing atmospheric drying (increases in VPD).

@@

\subsection{Functional Form of the Scaling Term}
\label{scale_term}
While the above discussion of the sign of $\frac{\partial \; ET}{\partial \; VPD}$ is important to answer our question of when ET reponse increases or decreases with VPD, understating the overall magnitude of the ET response is important to soil-plant-atmosphere water budgeting. So we now more closely examine the terms that affect how the sign term is scaled:

@@

$\Delta$'s presence in the scaling term also matches physical intuition. $\Delta$ (and also the approxmately constant $\gamma$) control the efficiencies with which surface energy is converted to latent and sensible heat \citep{Monteith_1965}. The functional from of $\Delta$ will be the same across PFTs, but the temperature range may vary slightly. In contrast, aerodynamic conductance will vary strongly with PFT due to the importance of surface roughness for aerodynamic conductance. So most of the differences in scaling between PFT should be in the aerodynamic conductance term.

@@

To summarize, variability between PFTs is mostly controlled by systematic differences in aerodynamic conductance, due to differences in surface roughness between each PFT, and posibly to a lesser extent wind conditions. In contrast, variability within PFT is also controlled by temperature, through Clausius-Clapeyron. But, aerodynamic conductance variability generally impacts the scaling term more than temperature, even within PFT.

@@

In this section we consider direct observations of ET response with eddy-covariance data, while including uncertainty with the $\sigma$ term (Section \ref{methods}). These observational results of ET response largely confirm our theoretical analysis, presented above (Table \ref{stats}). For all PFTs, mean ET response to increasing VPD is negative. However, ET response evaluated at the average of all variables (e.g. $\sigma$, $T$, $c_a$, $VPD$) is postive for CRO, and negative for all other PFTs. This difference in mean ET response as compared to the ET response at mean environmental conditions is due to the non-linear nature of the response, in which negative responses are generally larger magnitude than positive responses (Figure \ref{idealized_sign}). Therefore, both the mean ET response as well as the ET response at mean environmental conditions matches our expectations from the theory (Section \ref{sign_term}), with the exception that CRO observations are shifted more towards a negative ET response than we expect.

@@

Regarding the frequency of negative and positive ET response, all PFTs exibit a  decreasing ET response with increasing VPD (physiologically controlled, water conservative response) for the majority of observations. When listed by increasing frequency of negative response, the PFTs are generally ordered as we would expect them to be, from most water intesive to most water conservative: CRO (53\%), DBF (61\%), EBF (67\%), GRA (69\%), ENF (74\%), and CSH (97\%).

@@

\subsection{Validation of theory at eddy-covariance sites}
\label{testing}
We now compare more sophisticated distributions of the observed response to our simpified theory (Section \ref{sign_term}). Figure \ref{test_sign} shows the observed distribution of the sign term, as compared to what the theory would predict. Our goal was to capture the leading order behavior of the ET dependence on VPD. Given the assumptions we have made, and the uncertainties of flux tower observations themselves, we expect a relatively large amount of noise when reproducing the derivatives of ET. However, the data largely reproduces our theoretical analysis.

@@

For DBF our theory is successful towards this goal: the theory matches the leading order behavior of the function when uncertainty is included, and the observations match the theory with the addtion of noise. The VPD dependence, given by the slope parameter ($g_1$), follows the median values of each bin. Perhaps most importantly, the x-intercept, and thus VPD$_{crit}$, matches nearly exactly between the theory and the observations. Therefore the sign of the ET response to increases in VPD should be well matched, subject to the unavoidable constraints of noise, much of which comes from the observations themselves. The uncertainty is non negligible; there are many observations in each bin for which the the sign of observation is opposite the response predicted by the theory, but to leading order our theory matches the observations well.

@@

As Table \ref{stats} suggested, the theory does not match as well for CRO as for DBF and CSH. At low VPD the VPD dependence (due to $g_1$) is similar to the leading order behavior of the observations, but the theory's curve is shifted towards more positive values relative to the observations, likely due to some bias in uWUE. Most strikingly, the observations show a tendency to move back towards negative values at high VPD, which given the functional form of VPD dependence, is behavior that is impossible for the theory to capture. This general positive bias of the theory explains the shift towards negative ET response observed in the bulk statistics in Table \ref{stats}. The divergence between the theory and the observations for CRO could have been expected. For practical reasons, our theory does not include important sources of plant type heterogeneity within and across sites for CRO. In agricultural landscape, the footprint of an eddy-covariance tend to not only observe a single species but many differnent ones, with potentially varying water use strategies. We also do not account for seasonal changes in LAI, variability in crop height and surface roughness, differences in C3 vs C4 photosynthesis and water strategies, or differences in irrigation practices between the sites and years. These deficiencies should largely explain the inability of our theory to exactly match the observations in croplands. For example, a superposition of sites with C3 plants (at low environmental VPD) and C4 plants (at high environmental VPD) would explain observed shift back towards negative ET response at high VPD when all sites are binned together, as in Figure \ref{test_sign}.

The theory for GRA suffers from similar limitations as for CRO. GRA observations are characterized by a consistent trend back towards negative ET response at higher VPD, which the functional form of our theory is incapable of accounting for. As compared to CRO, the divergence between the theory and the observations is even greater for GRA, but these differences are explainable using similiar logic as we did for CRO. GRA also tends to exhibit substantial within-PFT variability, such as C3 vs C4 photosynthesis, as well as strong seasonal variability in height and LAI. As with CRO, the trend back towards negative ET response at higher VPD could be explained by a superposition of sites with C3 vs C4 sites. Unfortunately, because detailed site-level data of C3 vs C4 plant type is not available, we could not bin CRO and GRA further by irrigation and photosynthesis strategy. However, we hypothesize that the theory would validate against observations much better when photosynthesis pathways are accounted for.

The comparison between the theory and observations for ENF and EBF is unique from the rest of the PFTs. The VPD dependence of the theory, determined by the slope parameter ($g_1$), is not as well captured by observations. However, the theory still captures the general behavior and magnitude of the response, with some increase in error at low and/or high VPD.  We note that we used published constants for $g_1$ \citep[from ][]{Lin_2015, Franks_2017}, which determines the VPD dependence, and that deviations should be expected across sites. We could have fit our own constants so that the theory would best match the observations, but our goal was to use a priori results whenever possible in order to avoid overfitting and to have a theoretical approach as independent to the observations as possible for further confirmation of the validity of the approach.

@@

Most of the results presented so far focus on the sign term, so now we turn to observations of ET response with the scaling term included (Figure \ref{data_scatter}). Until now, in the interest of developing leading order intuition for ET response, and to be conservative in our acknowledement of model and observational error, we've considered $\sigma$ variability to be a measure of uncertainty. An alternative viewpoint is that $\sigma \cdot uWUE$ represents spatial and temporal variability of uWUE, which may be expected within bounds \citep[see Table \ref{pft}, also ][]{Zhou_2015}. This is a less conservative view; some of the $\sigma \cdot uWUE$ variability will be due to model and observational error so by viewing $\sigma \cdot uWUE$ as \textit{real} variability we run the risk of mistaking noise for signal. However, the advantage of this viewpoint is that, from \citet{Zhou_2014}, we have very high confidence that uWUE fit at the hourly timescale (as we do with $\sigma \cdot uWUE$) correctly captures the relationship between ET, GPP, and VPD, and that our form of PM introduces minimal error with its use of uWUE.

@@

Therefore, with the caveat that some of the signal presented in Figure \ref{data_scatter} may in fact be noise, we can interpret the observed distribution of ET response to VPD. In general, the observed response matches the intuitive theory. ET response to VPD shifts towards positive values as VPD increases (atmospheric demand dominating). CRO exhibit the highest occurence of postive ET response, and the obsevations confirm that CRO are the most water intensive. CSH are the most water conservative, with a strong negative response. DBF, EBF, ENF, and GRA are also water conservative, but show some occurence of positive ET response to VPD, particularly at higher observed VPD, as the theory predicts.

@@

As in Figure \ref{test_sign}, Figure \ref{data_scatter} matches our expectation based on simplified theory. The sign term is most strongly scaled by $g_a$, and in general the sign term shifts towards positive values as VPD observations increase. The willingness with which a given plant type evolved to use water dictates the occurence of positve versus negative ET response. Water conservative plants are highly effective at mitigating the effects of atmospheric demand, and can store water for later use by reducing ET in response to increasing VPD.

@@

Our theory suggests that for a majority of enviornmental conditions, plants will tend to conserve water and reduce ET with increasing VPD. Stomatal regulation and plant physiological response strongly regulate ET, and this regulation varies by PFT. CROs are the least water conservative, while DBF, EBF, ENF, GRA and CSH are progressively more water conservative (more likely to reudce ET in response to increasing VPD). Observations of ET response to VPD exhibit the same general behavior as the theory, with ET response becoming more postive (atmospheric demand dominating) as environmental VPD increases within a PFT, and more negative for PFTs that are adapted to arid conditions and prioritize water conservation over primary production.

The consequences of our theory our twofold. First, the ET response to VPD is always far more negative than it would be if only atmospheric demand is considered and plant phsiological reponse is ignored (as with PET). Therefore,  any drought indices or analysis using PET will not capture how  vegetated surfaces respond and react to increases in atmospheric demand. However, the new uWUE-version of Penman Monteith we derived (Equation \ref{et}) could be used as a complement to PET in drought indices. One could calculate and compare indices  when plant response is ignored (PET) or included (Equation \ref{et}).

Lastly, our paper builds important intuition for how plants will respond to global change-induced shifts in VPD, which should increase in the future. Plant physiological responses to direct CO$_2$ effects \citep[e.g.,][]{Swann_2016} receives more attention than physiological response to indirect effects like increased VPD. Here, we provide broad PFT-focused results showing a likely decrease in ET in response to positive VPD perturbations (amtopsheric drying). Addionally, we derive an explicit conceptual framework for ET response as a function of two plant type parameters, $g_1$ and uWUE (e.g. Equation \ref{et}). This framework can be used to understand ET reponse to a range of environmental change, and forms a foundation for resolving varying respose to more nuanced consideration of plant type and climates. Any plant physiological heterogeneity that can be conceptualized with shifts in $g_1$ \citep[e.g.][]{Lin_2015} and/or uWUE \citep[e.g.][]{Zhou_2014} is represntable within our framework, which opens the door for a heirarchy of more sophisticated climate- and plant-specific analysis of ET sensitivity to environmental variables (including VPD). We argue that such simplified conceptual frameworks are critical tools for disentangling land-atmosphere feedbacks at various scales, from diurnal to seasonal and beyond, and to characterize ET response in a warmer, atmospherically drier, and enriched CO$_2$ world.

>>>>

The opposing effects of increased atmospheric demand and higher stomatal stress lead to two possible perspectives for how evapotranspiration (ET) responds to shifts in VPD. The first, a hydrometeorological perspective, is that higher VPD increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indices and hydrometeorological studies \citep[e.g.,][]{Heim_2002, Scheff_2015}, typically only quantifies changes in atmospheric demand and fails to account for ecosystem response \citep{Swann_2016}. In reality, plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to partially close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. This leads to a plant physiology perspective, in which an increase in VPD, particularly in well-watered soil conditions, may actually correspond to a decrease in ET because of stomatal closure. This decrease in ET would reduce and potentially cancel, in the case of full stomatal closure, the effects of shifts to atmospheric demand. In other words, the  question ``When does VPD drive or reduce ET?'' can be related to whether plant regulation or atmospheric demand dominates ET response.

@@

\section{Materials and Methods}
\label{methods}
\subsection{Data}
\label{data}
We use both meteorological and eddy-covariance data from the FLUXNET2015 database, including all sites with at least four years of data (data available at \sloppy https://fluxnet.fluxdata.org/data/fluxnet2015-dataset/ \sloppy). Each site's plant functional type (PFT) was classified using the International Geosphere-Biosphere Programme vegetation classification scheme \citep{Loveland_1999}. The physical constants used in the stomatal conductance model in the Methods section (Section \ref{methods}) are only published for six plant functional types (PFTs): crops (CRO), deciduous broadleaf forest (DBF), evergreen broadleaf forests (EBF), evergreen needleleaf forest (ENF), grass (GRA), and closed shrub (CSH) (see Table \ref{pft}). There are 62 sites with these plant functional types, and their location is shown in  Figure \ref{map_fig}.

\begin{table}
  \caption{Plant functional types, their abbreviation, Medlyn coefficient (from \citet{Lin_2015}, via \citet{Franks_2017}) and calculated uWUE. uWUE from \citet{Zhou_2015}, including the  observed standard deviation is shown for comparison. Note that uWUE from \citet{Zhou_2015} is calculated from a different set of sites, and that units are converted such that the quantities fit into Equations 1-8 with the variables in Table \ref{definitions}.}
    \small
\label{pft}
\centering
\begin{tabular}{l c c @{\qquad} c c}
  \hline
  \multirow{2}[3]{*}{Abbreviation} & \multirow{2}[3]{*}{PFT} & \multirow{2}[3]{*}{$g_1$ (Pa$^{0.5}$)} & \multicolumn{2}{c}{uWUE ($\mu$-mol [C] Pa$^{0.5}$ J$^{-1}$ [ET])}  \\
  \cmidrule{4-5}

@@

Differences in $VPD_{crit}$ are exclusively determined by uWUE and the slope parameter ($g_1$) related to the plant functional type. A larger uWUE means a smaller $VPD_{crit}$, and an ET response to increases VPD that is more likely to be positive. At first glance this result is somewhat counter-intuitive; we expect that plants with a higher water use efficiency would be more water conservative. However, in reality uWUE determines how $WUE$ changes with VPD:

@@

So, plants with a higher uWUE will have a greater \textit{decrease} in ecosystem-scale $WUE$ in response to increases in VPD. This decrease in $WUE$ causes more water loss per unit carbon production, and explains the relationship between high uWUE and high likelihood of increases of ET in response to increasing atmospheric drying (increases in VPD).

@@

\subsection{Functional Form of the Scaling Term}
\label{scale_term}
While the above discussion of the sign of $\frac{\partial \; ET}{\partial \; VPD}$ is important to answer our question of when ET response increases or decreases with VPD, understating the overall magnitude of the ET response is important to soil-plant-atmosphere water budgeting. So we now more closely examine the terms that affect how the sign term is scaled:

@@

$\Delta$'s presence in the scaling term also matches physical intuition. $\Delta$ (and also the approximately constant $\gamma$) control the efficiencies with which surface energy is converted to latent and sensible heat \citep{Monteith_1965}. The functional from of $\Delta$ will be the same across PFTs, but the temperature range may vary slightly. In contrast, aerodynamic conductance will vary strongly with PFT due to the importance of surface roughness for aerodynamic conductance. So most of the differences in scaling between PFT should be in the aerodynamic conductance term.

@@

To summarize, variability between PFTs is mostly controlled by systematic differences in aerodynamic conductance, due to differences in surface roughness between each PFT, and possibly to a lesser extent wind conditions. In contrast, variability within PFT is also controlled by temperature, through Clausius-Clapeyron. But, aerodynamic conductance variability generally impacts the scaling term more than temperature, even within PFT.

@@

In this section we consider direct observations of ET response with eddy-covariance data, while including uncertainty with the $\sigma$ term (Section \ref{methods}). These observational results of ET response largely confirm our theoretical analysis, presented above (Table \ref{stats}). For all PFTs, mean ET response to increasing VPD is negative. However, ET response evaluated at the average of all variables (e.g. $\sigma$, $T$, $c_a$, $VPD$) is positive for CRO, and negative for all other PFTs. This difference in mean ET response as compared to the ET response at mean environmental conditions is due to the non-linear nature of the response, in which negative responses are generally larger magnitude than positive responses (Figure \ref{idealized_sign}). Therefore, both the mean ET response as well as the ET response at mean environmental conditions matches our expectations from the theory (Section \ref{sign_term}), with the exception that CRO observations are shifted more towards a negative ET response than we expect.

@@

Regarding the frequency of negative and positive ET response, all PFTs exhibit a  decreasing ET response with increasing VPD (physiologically controlled, water conservative response) for the majority of observations. When listed by increasing frequency of negative response, the PFTs are generally ordered as we would expect them to be, from most water intensive to most water conservative: CRO (53\%), DBF (61\%), EBF (67\%), GRA (69\%), ENF (74\%), and CSH (97\%).

@@

\subsection{Validation of theory at eddy-covariance sites}
\label{testing}
We now compare more sophisticated distributions of the observed response to our simplified theory (Section \ref{sign_term}). Figure \ref{test_sign} shows the observed distribution of the sign term, as compared to what the theory would predict. Our goal was to capture the leading order behavior of the ET dependence on VPD. Given the assumptions we have made, and the uncertainties of flux tower observations themselves, we expect a relatively large amount of noise when reproducing the derivatives of ET. However, the data largely reproduces our theoretical analysis.

@@

For DBF our theory is successful towards this goal: the theory matches the leading order behavior of the function when uncertainty is included, and the observations match the theory with the addition of noise. The VPD dependence, given by the slope parameter ($g_1$), follows the median values of each bin. Perhaps most importantly, the x-intercept, and thus VPD$_{crit}$, matches nearly exactly between the theory and the observations. Therefore the sign of the ET response to increases in VPD should be well matched, subject to the unavoidable constraints of noise, much of which comes from the observations themselves. The uncertainty is non negligible; there are many observations in each bin for which the the sign of observation is opposite the response predicted by the theory, but to leading order our theory matches the observations well.

@@

As Table \ref{stats} suggested, the theory does not match as well for CRO as for DBF and CSH. At low VPD the VPD dependence (due to $g_1$) is similar to the leading order behavior of the observations, but the theory's curve is shifted towards more positive values relative to the observations, likely due to some bias in uWUE. Most strikingly, the observations show a tendency to move back towards negative values at high VPD, which given the functional form of VPD dependence, is behavior that is impossible for the theory to capture. This general positive bias of the theory explains the shift towards negative ET response observed in the bulk statistics in Table \ref{stats}. The divergence between the theory and the observations for CRO could have been expected. For practical reasons, our theory does not include important sources of plant type heterogeneity within and across sites for CRO. In agricultural landscape, the footprint of an eddy-covariance tend to not only observe a single species but many different ones, with potentially varying water use strategies. We also do not account for seasonal changes in LAI, variability in crop height and surface roughness, differences in C3 vs C4 photosynthesis and water strategies, or differences in irrigation practices between the sites and years. These deficiencies should largely explain the inability of our theory to exactly match the observations in croplands. For example, a superposition of sites with C3 plants (at low environmental VPD) and C4 plants (at high environmental VPD) would explain observed shift back towards negative ET response at high VPD when all sites are binned together, as in Figure \ref{test_sign}.

The theory for GRA suffers from similar limitations as for CRO. GRA observations are characterized by a consistent trend back towards negative ET response at higher VPD, which the functional form of our theory is incapable of accounting for. As compared to CRO, the divergence between the theory and the observations is even greater for GRA, but these differences are explainable using similar logic as we did for CRO. GRA also tends to exhibit substantial within-PFT variability, such as C3 vs C4 photosynthesis, as well as strong seasonal variability in height and LAI. As with CRO, the trend back towards negative ET response at higher VPD could be explained by a superposition of sites with C3 vs C4 sites. Unfortunately, because detailed site-level data of C3 vs C4 plant type is not available, we could not bin CRO and GRA further by irrigation and photosynthesis strategy. However, we hypothesize that the theory would validate against observations much better when photosynthesis pathways are accounted for.

The comparison between the theory and observations for ENF and EBF is unique from the rest of the PFTs. The VPD dependence of the theory, determined by the slope parameter ($g_1$), is not as well captured by observations. However, the theory still captures the general behavior and magnitude of the response, with some increase in error at low and/or high VPD.  We note that we used published constants for $g_1$ \citep[from ][]{Lin_2015, Franks_2017}, which determines the VPD dependence, and that deviations should be expected across sites. We could have fit our own constants so that the theory would best match the observations, but our goal was to use a priori results whenever possible in order to avoid over-fitting and to have a theoretical approach as independent to the observations as possible for further confirmation of the validity of the approach.

@@

Most of the results presented so far focus on the sign term, so now we turn to observations of ET response with the scaling term included (Figure \ref{data_scatter}). Until now, in the interest of developing leading order intuition for ET response, and to be conservative in our acknowledgment of model and observational error, we've considered $\sigma$ variability to be a measure of uncertainty. An alternative viewpoint is that $\sigma \cdot uWUE$ represents spatial and temporal variability of uWUE, which may be expected within bounds \citep[see Table \ref{pft}, also ][]{Zhou_2015}. This is a less conservative view; some of the $\sigma \cdot uWUE$ variability will be due to model and observational error so by viewing $\sigma \cdot uWUE$ as \textit{real} variability we run the risk of mistaking noise for signal. However, the advantage of this viewpoint is that, from \citet{Zhou_2014}, we have very high confidence that uWUE fit at the hourly timescale (as we do with $\sigma \cdot uWUE$) correctly captures the relationship between ET, GPP, and VPD, and that our form of PM introduces minimal error with its use of uWUE.

@@

Therefore, with the caveat that some of the signal presented in Figure \ref{data_scatter} may in fact be noise, we can interpret the observed distribution of ET response to VPD. In general, the observed response matches the intuitive theory. ET response to VPD shifts towards positive values as VPD increases (atmospheric demand dominating). CRO exhibit the highest occurrence of positive ET response, and the observations confirm that CRO are the most water intensive. CSH are the most water conservative, with a strong negative response. DBF, EBF, ENF, and GRA are also water conservative, but show some occurrence of positive ET response to VPD, particularly at higher observed VPD, as the theory predicts.

@@

As in Figure \ref{test_sign}, Figure \ref{data_scatter} matches our expectation based on simplified theory. The sign term is most strongly scaled by $g_a$, and in general the sign term shifts towards positive values as VPD observations increase. The willingness with which a given plant type evolved to use water dictates the occurrence of positive versus negative ET response. Water conservative plants are highly effective at mitigating the effects of atmospheric demand, and can store water for later use by reducing ET in response to increasing VPD.

@@

Our theory suggests that for a majority of environmental conditions, plants will tend to conserve water and reduce ET with increasing VPD. Stomatal regulation and plant physiological response strongly regulate ET, and this regulation varies by PFT. CROs are the least water conservative, while DBF, EBF, ENF, GRA and CSH are progressively more water conservative (more likely to reduce ET in response to increasing VPD). Observations of ET response to VPD exhibit the same general behavior as the theory, with ET response becoming more positive (atmospheric demand dominating) as environmental VPD increases within a PFT, and more negative for PFTs that are adapted to arid conditions and prioritize water conservation over primary production.

The consequences of our theory our twofold. First, the ET response to VPD is always far more negative than it would be if only atmospheric demand is considered and plant physiological response is ignored (as with PET). Therefore,  any drought indices or analysis using PET will not capture how  vegetated surfaces respond and react to increases in atmospheric demand. However, the new uWUE-version of Penman Monteith we derived (Equation \ref{et}) could be used as a complement to PET in drought indices. One could calculate and compare indices  when plant response is ignored (PET) or included (Equation \ref{et}).

Lastly, our paper builds important intuition for how plants will respond to global change-induced shifts in VPD, which should increase in the future. Plant physiological responses to direct CO$_2$ effects \citep[e.g.,][]{Swann_2016} receives more attention than physiological response to indirect effects like increased VPD. Here, we provide broad PFT-focused results showing a likely decrease in ET in response to positive VPD perturbations (atmospheric drying). Additionally, we derive an explicit conceptual framework for ET response as a function of two plant type parameters, $g_1$ and uWUE (e.g. Equation \ref{et}). This framework can be used to understand ET response to a range of environmental change, and forms a foundation for resolving varying response to more nuanced consideration of plant type and climates. Any plant physiological heterogeneity that can be conceptualized with shifts in $g_1$ \citep[e.g.][]{Lin_2015} and/or uWUE \citep[e.g.][]{Zhou_2014} is representable within our framework, which opens the door for a hierarchy of more sophisticated climate- and plant-specific analysis of ET sensitivity to environmental variables (including VPD). We argue that such simplified conceptual frameworks are critical tools for disentangling land-atmosphere feedbacks at various scales, from diurnal to seasonal and beyond, and to characterize ET response in a warmer, atmospherically drier, and enriched CO$_2$ world.
